---
title: "Command Line, Shell, Scripting"
subtitle: "Software Carpentry Notes"
author:
 - name: Paul Johnson
   affiliation: Center for Research Methods and Data Analysis, University of Kansas
   email: pauljohn@ku.edu
 - name: "CRMDA Guide #41. Please visit [http://crmda.ku.edu/guides](http://crmda.ku.edu/guides) for updates." 
 - name: "Tags: command line, shell, Linux, Terminal"

abstract: 
    This uses knitr to interleave shell commands and output, using the CRMDA style sheet.
checked_by: "First Last"
Note to Authors: please_dont_change_the_next 4 lines!			
date: "`r format(Sys.time(), '%Y %B %d')`"
output:
  html_document:
    highlight: haddock
---

```{r setup, include=FALSE}
##This Invisible Chunk is required in all CRMDA documents
outdir <- paste0("tmpout")
options("prompt" = "$ ")
options("continue" = "$ ")

if (!file.exists(outdir)) dir.create(outdir, recursive = TRUE)
knitr::opts_chunk$set(engine="bash", echo=TRUE, comment="output:", prompt = TRUE, fig.path=paste0(outdir, "/p-"))
options(width = 70)
```

## Getting Started

### Start a Terminal Emulator
   * Mac (Utilities/Terminal, iTerm, many others)
   * Windows (Git for Windows provides a BASH shell)
   * Linux (many terminal programs, some more exciting than others)

### High points
   * Type commands, see textual output
   * Not much pointing and clicking (copy/paste possible on some)
   
### What is a Shell?
   * Computer has many settings that programs can access. A high level program called a "shell" keeps those settings and passes them out to programs when desired
   * BASH: Bourne Again Shell, a widely-used shell
   * A terminal program relies on the shell to translate between user and operating system
   
#### **Prompt**. Where you type {.bs-callout .bs-callout-info}
   * Prompt might be verbose 
```
pauljohn:Documents/Projects $
```
   * On some it is very lean, just the dollar sign
```
$
```
   * Dollar sign is customary prompt for non-root user (non-administrators)
   * Prompt is configurable, it is a good exercise for somebody who has used the Terminal for a week or two on a daily basis.
   
## Basic things to type

#### What is my user name? Run "whoami" {.bs-callout .bs-callout-warning}
```{r whoami}
whoami
```
   * What's that doing? *It asks **the environment** for the value of* "$USER". 
   * We should get the same if we run
```{r whoami2}
echo $USER
```
**echo**: prints out a value to the screen
   
   * I have usually done that, rather than `whoami`.
   * To be perfectly honest, I did not know `whoami` existed until I watched an SC video. 


#### Why did USER have a dollar sign on it?

    * In the standard shell, variables in *the environment* are named with capitals, like "USER"
    * but to access (retrieve) we need to add dollar sign

#### Where Am I? Run "pwd"

```{r where0, eval=F, include=T}
pwd
```
**pwd**: print working directory

   * Start a "fresh" terminal, the output for the working directory is *almost always* the user HOME folder.

#### Fastest way to HOME folder
```{r}
cd
```
    * Will see later that cd is a multipurpose command.

#### Where is pauljohn as he works on this document? 
```{r where1, include=T}
pwd
```

#### Where's Home? Operating Systems Differ {.bs-callout .bs-callout-warning}
```{r showhome}
echo $HOME
```

   * Linux  "/home/pauljohn"
   * Macintosh  "/Users/pauljohn"
   * Windoze  "/c/Users/pauljohn"
       * Even though Windows refers to its directory as "C:\\Users\\pauljohn", BASH won't!
   * FORWARD SLASHES separate directory and file names
   * Shell "homogenizes" appearance of directories and files across operating systems, making them look Unix-like.

#### What is especially important about HOME? {.bs-callout .bs-callout-danger}
   * HOME is the only folder where a user is sure to be able to read and write
   * HOME is where the configuration files are stored. 
   * Customary: config files and directories begin with the letter "."
   * Many file-listing programs will NOT print the items that begin with "."
   * In my HOME, I have folders like ".mozilla", ".thunderbird", ".config", etc
   * $HOME is often referred to by a shortcut symbol, "~" (tilde)
   
#### Setting HOME {.bs-callout .bs-callout-warning} 
   * Usually, when you start a terminal program, the working directory will be HOME.
   * Unix, Linux, Mac systems all use environment variable HOME as a key organizing concept.
   * Windows does not, by default, have an environment variable named HOME!
       * Some Terminals or Shells will try to guess what HOME ought to be
       * Windows has other elements in the environment which programs use to guess the HOME
   * If HOME is not correct, as sometimes happens on Windows
      * Must fix ASAP, or else nothing we do in the shell will make sense!
   * Know what HOME ought to be? Can set manually. 
```
export HOME="/c/Users/pauljohn"
```
    * But there is a way that usually gets the correct answer on Windows, where this is the only place that we see this problem:

```
export HOME=%USERPROFILE%
```
**export**: A BASH shell command that takes the following value and places it into the environment for the duration of the session.

### Portable concepts: All Terminals Today will have

   * History. 
     * Usually accessed by up arrow key
     * or the command `history`
   * **Tab completion**: start typing a command or directory or file name and hit the TAB key
   * Optional color coded output of file and directory listings

# Listings

#### The "ls" function is a workhorse {.bs-callout .bs-callout-warning} 

   * ls is a short name because it is used very often
```{r, eval=F}
ls
```
   * Lets make sure your terminal has color-coded output
      * files should be visibly different from directories or other items
```{r, eval=F}
ls --color=always
```

   * "--color=always" is a *long form* command line **argument**. 
   * Short form arguments with `ls` are `-l`, `-a`, `-S`, `-1`, `-F`
   
   * See why color-coded output in the Terminal was a major breakthrough?
```{r, eval=F}
ls --color=never
```
   * In the olden days, before color-coded output, they would add the argument -F to put a slash on the end of directory names so that they were visibly different
```{r}
ls --color=never -F 
```

### The ls arguments I use most often are `l` and `a`

    * Default will not list "hidden" items, things that begin with period
```{r}
cd
ls -a
```

   * The -l argument asks for more comprehensive information
```{r}
ls -l
```

   * Often I put the two together
```{r}
ls -la
```

   * The far left part is read, write, and execute permissions.
   * Permissions and user group concepts differ across Operating systems and the BASH shell's view of them is not entirely accurate all the time
      * Well, its accurate, but not "fine-grained".

#### And Now, some PATH magic {.bs-callout .bs-callout-warning} 

   * We can repeatedly display contents of subdirectories
      * Run `ls`, hopefully you see at least one directory. Suppose that's called "Documents"
      * Try this
```{r, eval=F}
ls Documents
```

      * If we are lucky, Documents has another directory, say "Papers" inside it

```{r, eval=F}
ls Documents/Papers
```

```{r, eval=F}
ls Documents/Papers
```






```{r test-bash, engine='bash'}
echo hello world
echo 'a b c' | sed 's/ /\|/g'
```

How about `sh`?

```{r test-sh, engine='sh'}
# run wc on all engine examples
ls | grep engine | head -n8 | xargs wc
```








## In our KU css file, we have special highlighted elements for callouts.  

Callout usage is demonstrated in the following. These
   are always preceeded by four # signs. Yes, ####. We stole this idea
   from other styles and put them into ours. See usage of {.bs-callout
   .bs-callout-whatever_you_put here}, where whatever_you_put_here is
   either "danger", "warning" or "info". We might be able to negotiate
   you you on the choice of colors.  Run the render script with this
   document, you'll see what we mean.
	

#### Danger: Elements to fix {.bs-callout .bs-callout-danger}
1) topical tags, first of which is "guide"

      a) we need a list of legal tags

      b) way to enforce the tags

      c) way to make them available better than making them a co-author

2) Wish header had comment like ("this document was retrieved from
CRMDA") without making it a coauthor

#### Note: Policies about writing in these documents. {.bs-callout .bs-callout-warning}

1. Blank rows separate paragraphs.

2. The character width of rows should be 80 or less.  I have
no idea how anybody thinks they have a right to impose an
infinitely long row, but its bad.  Edit the document
with Emacs, run M-q to get repositioned text.  If your
editor cannot do that, quit using it.

#### Info: Use these callouts to attract attention {.bs-callout .bs-callout-info}

1. You can use danger (red), warning (orange), or info (blue)

2. Must make sure compiling using the kutils.css located in the kutils package
e.g., `rmarkdown::render("template-guide.Rmd", html_document(css = system.file("rmarkdown/",`  
`"kutils.css",package = "kutils")))`

## You have the ability to include tabs {.tabset .tabset-fade}

If you'd like, you can include different tabs. Maybe you want to show two similar versions of the same analysis. Or maybe something else. Whatever it may be, be sure to include `{.tabset .tabset-fade}` along with a level 2 header (##). For each tab, put the information under the level 3 header (###). Use another level 2 header to get out of the tabs mode.

The only way (that we know of) to get colors is to wrap the tab headers in a `<span style>` as shown below. This might be useful to draw attention to the tabs. Blue is the default color.

#### Info: What it looks like {.bs-callout .bs-callout-info}

```
## You have the ability to include tabs {.tabset .tabset-fade}

If you'd like, you can include different tabs. Maybe you want to show two similar versions of the same analysis. Or maybe something else. Whatever it may be, be sure to include `{.tabset .tabset-fade}` along with a level 2 header (##). For each tab, put the information under the level 3 header (###). Use another level 2 header to get out of the tabs mode.

The only way (that we know of) to get colors is to wrap the tab headers in a `<span style>` as shown below. This might be useful to draw attention to the tabs. Blue is the default color.

### <span style="color:red">First Tab Here</span>

Some information you'd like to show.

### <span style="color:orange">Second Tab Here</span>

Some more information you'd like to show.

### Third Tab Here

Some final information you'd like to show.

## Equations can be written in LaTeX, either "inline" or "display" mode.
```

## {.tabset .tabset-fade}

### <span style="color:red" "background-color:blue">First Tab Here</span>

Some information you'd like to show.

### <span style="color:orange">Second Tab Here</span>

Some more information you'd like to show.

### Third Tab Here

Some final information you'd like to show.

## Equations can be written in LaTeX, either "inline" or "display" mode.

Equations can be introduced in the style of LaTeX.  If the document
is processed into a pdf file, the LaTeX programs will be used, but
if the document is rendered to HTML, then symbolic questions are positioned
into the output and then the MathJax javascript framework is used to
display the math in the browser.

Typing the markup is like writing in LaTeX.  Two dollar signs ($\$$)
surround an inline mathematical expression while a display equation
(centered and separated) is bracketed by two dollar signs ($\$\$$).
The two dollar signs are often written on a line by themseves
for visual separation. We note that other methods that are used to
initiate display equations in latex also work, including $\[$ or
$\begin{equation}$.

\textbackslash{}begin\{equation\} and \textbackslash{}end\{equation\}
and \textbackslash{}[ and \textbackslash{}] generate the same. We need
to know if there is a recommended style.

Fractions (\\frac{}{}): $\frac{2}{3}$.

Hats (\\hat{}): $\hat{\lambda}=1.02$.

Square Root (\\sqrt{}), $\sqrt{4}=2$.

$$\alpha, \beta,  \gamma, \Gamma$$

Some popular binary operators are
$$ + - = \ne \ge \lt \pm $$
$$x \ge 15$$
$$a_i \ge 0~~~\forall i$$

## Matrix

$$
A_{m,n} =
 \begin{pmatrix}
  a_{1,1} & a_{1,2} & \cdots & a_{1,n} \\
  a_{2,1} & a_{2,2} & \cdots & a_{2,n} \\
  \vdots  & \vdots  & \ddots & \vdots  \\
  a_{m,1} & a_{m,2} & \cdots & a_{m,n}
 \end{pmatrix}
$$

This example uses the alternative display math $\[$, and
a different type of matrix.

\begin{equation}
A_{m,n} = \left(\begin{array}{ccc}
a & b & c\\
d & e & f\\
g & h & i
\end{array}\right)
\label{eq:another1}
\end{equation}

After some testing, it appears also that many
LaTeX symbols that we would expect to use have no benefit in Rmd
documents.



We don't know if there is a comprehensive list of which LaTeX math
symbols are allowed, but we wish somebody would find one.

